Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-15 talk/pepm2011/paper.tex
kill the section about frames, as michael thinks its too confusing. Instead mention frames minimally in the evaluation section.

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
objects that are allocated in the trace \emph{static,} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}; a more formal description is
given in Section~\ref{sec:formal}.

In Section~\ref{sec:frames} we describe some supporting techniques that are not
central to the approach, but are essential to improve the results. The introduced
techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
interpreter as a case study.

@@

The optimization is particularly effective for chains of operations.
For example, it is typical for an interpreter to generate sequences of
writes-followed-by-reads, where one interpreted opcode writes to some
object's field and the next interpreted opcode reads it back, possibly
dispatching on the type of the object created just before.  In the case
of PyPy's Python interpreter, this optimization can even remove the
allocation of all intermediate frames that occur in the interpreter,
corresponding to all calls that have been inlined in the trace.

@@

\section{Taming Frame Objects}
\label{sec:frames}

\cfbolz{probably can be cut in case of space problems}

XXX unclear what the relationship is to the technique; has it been used in the experiments??
XXX Maybe move to discussion and related work section?

One problem to the successful application of the allocation removal
techniques described in the previous sections is the presence of
frame-introspection features in many dynamic languages and the fact
that frames are heap allocated. Languages such as Python and Smalltalk
allow the programmer to get access to the frames object that the
interpreter uses to store local variables. This is a useful feature,
as it makes the implementation of a debugger possible in Python
without extra support from the VM. On the other
hand, it severely hinders the effectiveness of allocation removal,
because every time an object is stored into a local variable, it is
stored into the frame-object, which makes it escape.

This problem is solved by making it possible to the interpreter author
to add some hints into the source code to declare instances of one
class as frame objects needing special treatment. The JIT will then
fill these objects only lazily when they are actually accessed (\eg
because a debugger is used) using values from JIT-emitted code runtime
locations (typically the CPU stack). Therefore in the common case,
nothing is stored into the frame objects, making the problem of too
much escaping go away. This special handling of frame objects is a common approach in VM implementations
\cite{miranda_context_1999,andreas_gal_trace-based_2009}. The only novelty in our approach lays in
its generality: all the delicate support code for this is generated, as opposed to
most other JITs, which are just specifically written for one particular
language.

% section Taming Frame Objects (end)

\section{Evaluation}
\label{sec:Evaluation}

>>>>

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
objects that are allocated in the trace \emph{static,} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}; a more formal description is
given in Section~\ref{sec:formal}. The introduced
techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
interpreter as a case study.

@@

The optimization is particularly effective for chains of operations.
For example, it is typical for an interpreter to generate sequences of
writes-followed-by-reads, where one interpreted opcode writes to some
object's field and the next interpreted opcode reads it back, possibly
dispatching on the type of the object created just before.  A typical example
would be chains of arithmetic operations.

@@

\section{Implementation and Evaluation}
\label{sec:Evaluation}

The allocation removal techniques described in this paper were implemented in
the optimizer of PyPy's tracing JIT. The optimization is independent of which
interpreter a JIT is generated for. There are some practical issues beyond the
techniques described in this paper. The actual implementation needs to deal with
more operations than described in Section~\ref{sec:formal}, for example to
also support static arrays in addition to static objects. The implementation of
this optimization is about 400 lines of RPython code.

A further complication is that most interpreters written with PyPy use
heap-allocated frame objects to store local variables. Those severely hinders
the effectiveness of allocation removal, because every time an object is stored
into a local variable, it is stored into the frame-object, which makes it
escape. We implemented a general technique to treat such frames objects in a
special way to solve this problem. This is a common approach in VM
implementations \cite{miranda_context_1999,andreas_gal_trace-based_2009}. The
only novelty in our approach lays in its generality: most other JITs, are just
specifically written for one particular language instead of being generated from
an interpreter.
